# Chapter 1: Descriptive Statistics

## 1.1 Introduction to Descriptive Statistics

Descriptive statistics are methods used to summarize and describe the main features of a dataset. Unlike inferential statistics, which make predictions or inferences about a population based on a sample, descriptive statistics simply describe what the data shows.

### Key Purposes:
- **Summarize** large amounts of data in a meaningful way
- **Identify** patterns and trends
- **Communicate** findings clearly and concisely

## 1.2 Measures of Central Tendency

Central tendency refers to the center or typical value of a dataset. The three main measures are:

### Mean (Average)
The arithmetic mean is calculated by summing all values and dividing by the number of observations:

$$\bar{x} = \frac{\sum_{i=1}^{n} x_i}{n}$$

### Median
The middle value when data is ordered from smallest to largest.

### Mode
The most frequently occurring value in the dataset.

### Example in R

```{r central-tendency, fig.cap="Distribution of Student Exam Scores"}
# Sample data: exam scores
exam_scores <- c(78, 85, 92, 88, 76, 95, 88, 82, 90, 88, 79, 91, 87, 85, 88)

# Calculate measures of central tendency
mean_score <- mean(exam_scores)
median_score <- median(exam_scores)
mode_score <- as.numeric(names(sort(table(exam_scores), decreasing = TRUE)[1]))

# Display results
cat("Mean:", mean_score, "\n")
cat("Median:", median_score, "\n")
cat("Mode:", mode_score, "\n")

# Visualize the distribution
hist(exam_scores, 
     main = "Distribution of Exam Scores",
     xlab = "Score",
     ylab = "Frequency",
     col = "lightblue",
     border = "darkblue",
     breaks = 8)
abline(v = mean_score, col = "red", lwd = 2, lty = 2)
abline(v = median_score, col = "green", lwd = 2, lty = 2)
legend("topright", 
       legend = c("Mean", "Median"),
       col = c("red", "green"),
       lty = 2, lwd = 2)
```

## 1.3 Measures of Dispersion

Dispersion measures describe how spread out the data is.

### Range
The difference between the maximum and minimum values:

$$\text{Range} = \max(x) - \min(x)$$

### Variance
The average of squared deviations from the mean:

$$s^2 = \frac{\sum_{i=1}^{n} (x_i - \bar{x})^2}{n-1}$$

### Standard Deviation
The square root of variance, expressed in the same units as the data:

$$s = \sqrt{s^2}$$

### Example in R

```{r dispersion, fig.cap="Boxplot Showing Data Spread"}
# Calculate measures of dispersion
range_score <- max(exam_scores) - min(exam_scores)
variance_score <- var(exam_scores)
sd_score <- sd(exam_scores)

# Display results
cat("Range:", range_score, "\n")
cat("Variance:", variance_score, "\n")
cat("Standard Deviation:", sd_score, "\n")

# Create boxplot to visualize spread
boxplot(exam_scores,
        main = "Boxplot of Exam Scores",
        ylab = "Score",
        col = "lightgreen",
        border = "darkgreen",
        horizontal = FALSE)

# Add mean point
points(1, mean_score, col = "red", pch = 19, cex = 1.5)
legend("topright", 
       legend = "Mean",
       col = "red",
       pch = 19)
```

## 1.4 Comparing Multiple Groups

Let's compare exam scores across different study methods:

```{r group-comparison, fig.cap="Comparison of Study Methods"}
# Create data for three study groups
set.seed(123)
group_A <- c(78, 82, 85, 79, 88, 84, 81, 86, 83, 87)  # Traditional study
group_B <- c(88, 92, 90, 89, 95, 91, 93, 87, 94, 90)  # Group study
group_C <- c(72, 75, 78, 74, 80, 76, 73, 79, 77, 75)  # Minimal study

# Combine into a data frame
study_data <- data.frame(
  score = c(group_A, group_B, group_C),
  method = rep(c("Traditional", "Group Study", "Minimal"), each = 10)
)

# Summary statistics by group
aggregate(score ~ method, data = study_data, FUN = function(x) {
  c(Mean = mean(x), SD = sd(x), Median = median(x))
})

# Create comparative boxplot
boxplot(score ~ method, data = study_data,
        main = "Exam Scores by Study Method",
        xlab = "Study Method",
        ylab = "Score",
        col = c("lightblue", "lightgreen", "lightcoral"),
        border = c("darkblue", "darkgreen", "darkred"))
```

## 1.5 Summary

In this chapter, we covered:

- **Measures of Central Tendency**: Mean, median, and mode help us understand the "typical" value
- **Measures of Dispersion**: Range, variance, and standard deviation describe data spread
- **Visual Representations**: Histograms and boxplots provide intuitive understanding of distributions
- **Group Comparisons**: Comparing statistics across groups reveals patterns and differences

These foundational concepts are essential for understanding more advanced statistical methods in subsequent chapters.

---

## Navigation

- [← Back to Index](index.qmd)
- [Next: Chapter 2 →](02_chapter.qmd)
